Regime shifts and transformations in social-ecological systems: 

### 2025-09-29 review_paper/manuscript.Rmd
incorporating co-authors comments

---
title: "Regime shifts and transformations: "
subtitle: "Advancing critical frontiers for safe and just futures"
author: |
    | Juan C. Rocha et al.$^1$ 
    | \affil{$^1$Stockholm Resilience Centre, Stockholm University, Stockholm, Sweden 10691; email: \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}}
output:
  rtf_document:
  pdf_document:
    keep_tex: yes
    keep_md: yes
    latex_engine: xelatex
    #template: template.tex
    includes:
      in_header: header.tex
csl: /Users/juanrocha/Documents/styles/annual-reviews.csl
bibliography: refs.bib
documentclass: ar-1col-S2O
header-includes:
- \markboth{Rocha et al.}{Regime shifts and transformations} 
---

@@

## word counts
## Abstract 150
## Intro 697
## Method 957
## Results 621
## Discussion 1434
## Conclusion 267
## Total main 3976 + 150 abstract
## 
## Targets: Annual Reviews or Environment and Resources
## 9300w, 2 figs, 150 refs (27 pages in published volume)
## required sections:
## - summary points (max 8)
## - Future issues (max 8)
## - Terms and definitions: max 20, definitions of 20 words each

# Outline
# 
# Problem setting [1000w]
# 
# Definitions [1000w, 500w each]
# 
# History of the concepts [1000w]
# 
# Commonalities and differences [500w]
# 
# Current research challenges [1500w]
# 
# Multidimensionality
# 
# Observability and detection
# 
# Management: ethical stakes
# 
# Navigating the space of positive and negative tipping points [1000w]
# 
# Conclusions [500w]
# 
# Key insights (to be updated):
# 
# Clarify the difference and overlaps between transformations and regime shifts theories and the use of concepts such as tipping points
# 
# Highlighting frontiers of research that need both approaches
# 
# Navigating the space of positive and negative tipping points
# 
# \newpage
#

@@

\begin{abstract}
Current research problems in sustainability science require thinking on
nonlinear changes with central questions such as how to avoid them,
promote them, or predict them. Regime shifts and transformations
literature address some of these questions, yet there is little
cross-fertilization between these two streams of research. Here we
review them, propose avenues from future research and how together they
can be a powerful tool to achieve safe and just futures
\end{abstract}
\begin{keywords}
resilience, sustainability, transformations, adaptative capacity, regime shifts, adaptive management
\end{keywords}

@@

Sustainability scientists are concerned with nonlinear changes.
Depending on the research problem, the focus could be on avoiding undesirable changes, or promoting desirable ones.
But regardless of direction or desirability, notions of stability and resilience are common ground.
Examples of nonlinear changes include the adoption of new technologies or regulatory frameworks for e.g. the energy transition, the banning of damaging substances for human health or the environment (such as chlorofluorocarbons (CFC) and their impacts on the ozone layer), or the collapse of fish stocks.

There is two strands of sustainability science literature geared to deal with nonlinear changes that share a common origin [@holling1973resilience] but have been diverging in the last couple of decades: regime shifts [@folke2004] and transformations [@olsson2006shooting].
Regime shifts are large, abrupt and persistent changes in the function and structure of social-ecological systems [@biggs2018].
Transformations are about shifting development into new pathways, emergent and often unknown development trajectories [@folke2016].
We expand on definitions, similarities and differences of these two streams of work on the next section.
For now, we want to focus on why this review is timely and what we aim to bring to the larger scientific discussion.

Our review is motivated by the apparent separation of the two streams of work given that both come from the same conceptual framework [@holling1973resilience] but are applied in different ways by different scholarly audiences.
This could result in dimensions that are central on one framework becoming oversimplified in the other.
For example, agency —the ability of actors to influence each other and the environment— is central within transformations framing, but less prominent within regime shifts studies.
Conversely, the role of feedbacks is central in explaining regime shifts, but not always assessed in transformations research.

This lack of dialogue can lead to misuse of terminology, oversimplifciation, jargonification, or simply re-inventing the wheel.
For example, the concept of positive tipping points emphasize desirable transitions such as the adoption of electric vehicles [@lenton2025; @lenton2023global], but omit to analyze which societal groups will benefit or suffer from this transition and what it does imply for social dimensions of sustainability.
Such issues related to justice and power have been studied for three decades in transformations research [@pereira2024equity] and could have highlighted the societal risks entailed in energy transitions.

Sustainability science problems often bring questions of desired for whom, whose voices are heard in decision making, or dimensions of justice with e.g. future generations.
Research in both the transformations and regime shifts strains have already contributed important milestones and tools to deal with such questions.
However, these communities ignoring each other risks missing opportunities or the lack of cross-fertilization between the fields.
In other words, it reduces the probability that we can successfully solve some of the core problems of sustainability science, which by definition, require understanding social and biophysical dynamics simultaneously.

Sustainability science suffers from a proliferation of conceptual frameworks that grows faster than the capacity of scientist to test their validity or adequacy to explain real world problems.
It is not alone, similar scholarly dynamics have been reported in sociology, ecology or physics [@clark2020; @watts2014; @marquet2014].
A diversity of frameworks is an asset when a sustainability scientist treat them as tools to solve research problems.
But it becomes problematic when similar ideas are re-framed over and over under different semantics, yet none of them advance from framework to scientific theory, that is a set of causal statements that produce refutable expectations.

For example, the idea that social and ecological dynamics are intertwined, inseparable or entangled is at least three decades old [@berkes2000linking; @berkes2008navigating; @folke1996; @folke1991; @günther1993; @levin1999fragile; @holling2001; @gunderson2002panarchy], if not older [@renn2020evolution].
In fact, the distinction between the natural and social does not exist in many indigenous cosmologies.
Multiple frameworks have attempted to reshape the idea, from the social-ecological systems framework (SES) [@ostrom2009], socio-ecological technological systems (SETS) [@mcphearson2022; @ramaswami2012; @mcphearson2021], coupled human environmental systems (CHENS) [@liu2007] and more recently relational approaches [@mancillagarcia2024; @walsh2020], to name a few.
More matured frameworks such as the SES have hundreds of case studies coded supporting the internal and external validity of the framework, and suggesting under which conditions an SES is likely to succeed in sustaining e.g. common pool resources [@ostrom2007, e.g. [Library on Governance of Social-Ecological Systems](https://seslibrary.asu.edu)].
Other frameworks lack such rich empirical evidence, although some efforts are on their way in e.g. modeling transformations combining case studies and relational approaches [@schlüter2025].

As we will review, some of the criticisms of regime shifts and transformations are lack of comparative case studies that shed light under which conditions these frameworks have explanatory power, or under which conditions are they likely to fail.
An additional risk of multiple untested frameworks is the use of slightly different wording to mean the same thing (jargonification), for example positive tipping points and transformations.
Here we will review and clarify the nuance on meanings to avoid falling into the jargon trap, or at least make the content more accessible from a pedagogical point of view.

This review aims to outline research frontiers within sustainability science by looking closely to the transformations and regime shifts literature.
We first dive on the definitions and history of these concepts, trying to disentangle what are their common grounds and differences.
We then outline current research challenges and areas where we believe the two strands of work could benefit from working together.
Last, we conclude with a short agenda for future research.

@@

While we focus on the phenomena of regime shifts and transformations, the central character of the story is actually the object of study to which these phenomena can happen: social-ecological systems.
Social-ecological systems (SES) are human and natural coupled systems where people interact with biophysical components; they often exhibit nonlinear dynamics, feedbacks, time lags, heterogeneity and resilience [@liu2007a].
Resilience is the ability of a system to withstand disturbances without losing its function, structure and identity [@folke2016; @holling1973resilience].
The resilience scholars have adopted two important evolutionary features: resilience is also the capacity to adapt and transform.
Adaptability is the capacity to learn, combine experience and knowledge to adjust responses, while transformability is the ability to cross thresholds and move the SES into new basins of attraction, into emergent and often unknown development trajectories [@folke2016].

Historically, the conceptualization of resilience has its roots in mathematics.
Holling in his seminal work introducing the concept of resilience [@holling1973resilience] uses bifurcation theory (or catastrophe theory) to suggest minimal cartoons on how ecosystems can change, how they can lose stability and shifts from one configuration to another.
His work would have not been possible without previous developments on bifurcations (e.g. [@Poincare:1885ji; @thom1972stabilité] ).
Several concepts such as the basin of attraction (or attractor), tipping points, thresholds, or stability landscapes have precise definitions within these theories; but sometimes used as metaphors in later conceptualizations.
A basin of attraction, for example, is the region of parameter space that renders the same macroscopic qualitative behavior, it can be an equilibrium point, but also some more complex behaviour such as (a)periodic cycles, long transients, or chaos [@strogatz1994nonlinear].
We will not present formal definitions of these concepts, but instead direct the interested reader to introductory books on the matter [@strogatz1994nonlinear; @sole2011phase; @kuznetsov2023].

In the mathematical context, resilience is the size of the basin of attraction [@holling1973resilience; @ives1995; @Clark1975; @krakovská2023].
Features such as the distance to the boundary, the deepness or slope are often proposed as proxies of resilience [@Clark1975; @Mitra:2015jb; @Menck:2013jq] and enable the development of early warning systems for the proximity to tipping points [@scheffer2009].
Bifurcation theory, and thus notions of resilience, have been successfully applied to fields other than sustainability science and ecology, including but not limited to language, climate systems, the states of matter, cancer dynamics, finance, social dynamics such as poverty traps, believe spreading, voting behavior, segregation, the states of mind, or music, to name a few [@sole2011phase; @scheffer2009critical].
Some of these applications preceded the work of Holling in ecology [@holling1973resilience], and outlines the success of bifurcation theory across the sciences.
However, evolutionary dynamics of adaptations and transformations are not well defined within the seminal work on bifurcations.
While some formalizations have been used to explore for example the evolution of innovations [@kauffman1987; @kauffman2000; @sole2011phase], it remains an open problem to theorize true generative models that produce agents with behaviors that were not previously coded by the programmer [@kauffman2023].
This problem is common ground to both regime shifts and transformations research.

@@

Regime shifts are abrupt and persistent changes in the structure and function of social-ecological systems [@folke2004; @biggs2018].
They occur when resilience is eroded and the SES can easily depart from its original configuration and explore different regimes or modes of behavior.
Examples of regime shifts includes the transition from corals to algae dominated reefs [@hughes2017], eutrophication of lakes [@carpenter2020], the shift from forest to savannahs [@hirota2011; @staver2011], the disappearance of Arctic Sea Ice [@livina2013], the collapse of the thermohaline circulation [@stocker1997; @stommel1961], or the collapse of cooperation [@rand2014; @nowak2006].
Regime shifts are important because they often affect the flow of ecosystem services that people benefit from [@carpenter2009], and they present a difficult management problem since they can be hard to predict [@biggs2009; @hastings2010] and sometimes impossible to reverse on time scales relevant for management and policy [@biggs2009; @biggs2018].

@@

Regime shifts are typically characterized by their feedbacks, tipping points, thresholds, and hysteresis.
Feedbacks are mutually interacting processes that can act to amplify or damp its response to a force [@roe2009].
For example, the moisture recycling feedback is responsible for the maintenance of tropical rain forest like the Amazon, the more forest there are, the more humidity is transpired, which in turn feeds the clouds in the atmosphere that rains back to support the water demands of the forest [@zemp2017].
In drylands, facilitation feedbacks through shading, soil moisture and nutrient retention are responsible for shifts between grasslands, shrublands and deserts [@maestre2016; @berdugo2022].
In coral reefs, herbivory plays a fundamental role in mediating the shift between corals and macro-algae [@hughes2007].
Or in social systems, poverty traps are feedbacks mediated by income, education, or disease, that keeps people in poverty despite development efforts [@bowles2016poverty; @banerjee2012poor].
Balancing feedbacks are responsible for the stabilization of a particular regime, while reinforcing feedbacks are responsible for out of equilibrium dynamics [@biggs2018; @roe2009].
Thus weakening or strengthening of feedbacks can trigger nonlinear transitions.

@@

Tipping points are critical parameter values at which a system change its long term qualitative behavior.
They are bifurcation points or sigularities [@strogatz1994nonlinear; @thom1972stabilité].
Speaking of a point implies the assumption of a one-dimensional system with one co-dimension, that is a very simplified version of reality where there is only one state variable that we care about (e.g. one species) and one driver affecting it (e.g. temperature).
In reality, social-ecological system are complex adaptive systems composed by multiple interacting entities (e.g. species, businesses, cities) that are driven by multiple interacting forces that are not necessarily reducible to one (co)dimension [@levin1999fragile; @rocha2015].
When thinking of tipping phenomena in real life contexts, one should think on tipping lines or an hypersurface, that is all the combination of critical parameters of all potential drivers and state variables, rather than points.
Tipping points and thresholds are often used interchangeably.
However, in mathematics they have a slightly different meaning, the threshold is the separatrix between two basins of attraction, while the tipping point is the critical value at which a bifurcation occurs [@strogatz1994nonlinear].

Observing and measuring tipping points or thresholds in real life context is very difficult, and for that reason sometimes deemed not useful beyond theoretical work [@hillebrand2020; @kopp2024].
However, scientists have been able to identify some tipping points with different degrees of certainty.
For example, hypoxia and anoxia are regime shifts common in coastal areas where dissolved oxygen fall under levels that are not tolerable for different groups of organisms.
Levels below $2mlO_{2}/L$ gives rise to hypoxia a condition where many fish die-off, and below $0.5mlO_{2}/L$ to annoxia a condition that is toxic for humans [@diaz2008; @breitburg2018].
Tipping points in precipitation have been proposed for the transitions from forest to savanna ($1500mm$), savanna to forest ($2000mm$), and savanna to deserts ($<500mm$) [@hirota2011], which have been confirmed in observational studies [@staver2011; @aleman2020; @verbesselt2016].
Similarly, global thresholds for aridity has been reported for dryland ecosystems (0.54, 0.7 and 0.8) [@berdugo2020], while thresholds for deforestation in rain forest have been proposed (40% for the Amazon, 25% under precautionary principle, currently at 20%) [@nobre2016; @lovejoy2018].
Several Earth subsystems (e.g. the thermohalince circulation, the Greenland ice sheet) can have tipping points driven by temperature.
A recent expert elicitation speculates that six of these subsystems are highly likely to experience tipping dynamics under $2^\circ C$ of warming, with additional four tippping under $4^\circ C$ of warming [@armstrongmckay2022].

Hysteresis is the reason why some regime shifts are really hard, sometimes impossible to reverse.
Hysteresis means that tipping point in one direction of the shift is different from the tipping point in the opposite direction.
Thus one has to make additional efforts to restore social-ecological systems after they have tipped.
Much research has been devoted to understand how to manage regime shifts, how to prevent them, or what type of interventions are necessary and sufficient for recovery.
Instead of listing the different strategies that apply to different ecosystems and contexts (see the Regime Shifts Data Base), here we summarize an approach that has been successful in managing nonlinear dynamics in the face of uncertainty.

Adaptive management was developed and advocated by early regime shift scholars to encourage monitoring and experimentation, reduce uncertainty, and inform decision making [@holling1978adaptive; @walters1986adaptive].
The purpose of adaptive management is to proactively learn by experimentation.
It is widely use for managing fisheries, where monitoring efforts are combined with modeling to inform maximum sustainable yields in the near future; projections that are then updated through landing statistics and independent scientific monitoring [@walters1986adaptive].
Another example is fire management, where prescribed fires can maintain a structure of the landscape that provides memory to future disturbances and prevent synchronized fires [@peterson2002].
Experimentation in lakes have shown that they can be recovered from eutrophication by manipulating nutrient inputs, dredging, or biomanipulation of the food web [@carpenter1999; @smith2009; @scheffer1993].

@@

Another framework to think of transformations is positive tipping points [@lenton2025; @positive2024].
The concept has been widely used in climate sciences to mean conditions and strategies for fast decarbonization, including but not limited to, the adoption of new technologies such as electric vehicles, solar panels, the electrification of grids, or changing diets.
The literature on positive tipping points have been criticized for over emphasizing technocratic solutions while ignoring core issues of justice and distributions that transformation research have studied for at least three decades [@pereira2024equity].
Who defines what is positive, and positive for whom, are political questions often ignored.
The problem is that the solutions landscape risk to reinforce unjust and colonial practices of exploitation of minorities or less empowered.
It can lead to temporary or localized shifts (e.g. more electric cars) without fundamentally altering the underying system dynamics, values and norms that reinforced the undesired regime in first place [@milkoreit2022].

@@

>>>>

---
title: "Regime shifts and transformations in social-ecological systems: "
subtitle: "Advancing critical frontiers for safe and just futures"

author: |
    | \small Juan C. Rocha$^{1,2,§}$, Caroline Schill$^{1,3}$, Emilie A. L. Lindkvist$^{1}$, Reinette Biggs$^{1,4}$, Thorsten Blenckner$^{1}$, Anne-Sophie Crépin$^{1,3}$, Ingo Fetzer$^{1}$, Carl Folke$^{1,2,3}$, Amanda Jiménez-Aceituno$^{1,5}$, Jan J. Kuiper$^{1}$, Steven J. Lade$^{1,6}$, Carla Lanyon-Garrido$^{1,7}$, Romi Lotcheris$^{1}$, Romina Martin$^{1}$, Vanessa Masterson$^{1,8}$, Petr Matous$^{9}$, Michele-Lee Moore$^{1,10}$, Magnus Nyström$^{1}$, Per Olsson$^{1}$, Laura M. Pereira$^{1,11}$, Garry Peterson$^{1}$, André Pinto da Silva$^{1,12}$, Sasha Quahe$^{1}$, Maja Schlüter$^{1}$, Lan Wang-Erlandsson$^{1,2,13}$, and Hannah Zoller$^{1}$
    | 
    | \tiny $^{1}$Stockholm Resilience Centre, Stockholm University, 10691 Stockholm, Sweden
    | \tiny $^{2}$Anthropocene Laboratory, The Royal Swedish Academy of Sciences, 11418 Stockholm, Sweden
    | \tiny $^{3}$Beijer Institute of Ecological Economics, The Royal Swedish Academy of Sciences, 10405 Stockholm, Sweden
    | \tiny $^{4}$Centre for Sustainability Transitions (CST), Stellenbosch University, South Africa
    | \tiny $^{5}$Social-Ecological Systems Institute (SESI), Leuphana University of Lüneburg, Lüneburg, Germany
    | \tiny $^{6}$Fenner School of Environment & Society, Australian National University, Canberra, Australia
    | \tiny $^{7}$Energy Centre, Faculty of Physical and Mathematical Science, Chile University, Santiago, Chile
    | \tiny $^{8}$Anthropology Department, Rhodes University, Makhanda, South Africa
    | \tiny $^{9}$School of Project Management, University of Sydney, Sydney, Australia
    | \tiny $^{10}$Dept of Geography and Centre for Global Studies, University of Victoria, BC Canada
    | \tiny $^{11}$Global Change Institute, University of the Witwatersrand, Johannesburg, South Africa
    | \tiny $^{12}$cE3c — for Ecology, Evolution and Environmental Changes & CHANGE — Global Change and Sustainability Institute, Faculdade de Ciências, Universidade de Lisboa, Campo Grande, Lisbon, Portugal
    | \tiny $^{13}$Potsdam Institute for Climate Impact Research, Member of the Leibnitz Association, 14473 Potsdam, Germany
    | \tiny $^{§}$juan.rocha@su.se

output:
  pdf_document:
    keep_tex: yes
    keep_md: yes
    latex_engine: xelatex
    #template: template.tex
    includes:
      in_header: header.tex
  rtf_document:
csl: /Users/juanrocha/Documents/styles/annual-reviews.csl
bibliography: refs.bib
documentclass: ar-1col-S2O
header-includes:
- \markboth{Rocha et al.}{Regime shifts and transformations} 
- \graphicspath{{figures/}}
- \usepackage[switch]{lineno}
- \linenumbers
---

@@

dat <- googlesheets4::read_sheet(
    "https://docs.google.com/spreadsheets/d/1KmEF9qwiIn--WBDQULL_Plsu5Bw4RJrksnOatGFCQgU/edit?gid=0#gid=0", 
    sheet = 1, skip = 1, col_types = "ccccciiiiiiiiiiiiii"
) |> janitor::clean_names()

dat |> 
    rename(affiliation = 3, author_order = 6) |> 
    arrange(author_order, surname) |> 
    glue::glue_data("<<name>> <<surname>>$^{<<affiliation>>}$",
                    .open = "<<", .close = ">>") |> 
    str_flatten_comma(last = ", and ")

aff <- googlesheets4::read_sheet(
    "https://docs.google.com/spreadsheets/d/1KmEF9qwiIn--WBDQULL_Plsu5Bw4RJrksnOatGFCQgU/edit?gid=0#gid=0", 
    sheet = 2, skip = 0, col_types = "cc"
) |> janitor::clean_names()

glue::glue_data(aff, '    | \\tiny $^{<<code>>}$<<affiliation>>',
                .open = "<<", .close = ">>")

@@

\section*{Abstract}

\begin{abstract}
Current research challenges in sustainability science require us to consider nonlinear changes – shifts that do not happen gradually but can be sudden and difficult to predict. Central questions are therefore how we can prevent harmful shifts, promote desirable ones, and better anticipate both. The regime shifts and transformations literature is well-equipped to address these questions. Yet, even though both research streams stem from the same intellectual roots, they have developed along different paths, with limited exchange between the two, missing opportunities for cross-fertilisation. We here review the definitions and history of both research streams to disentangle common grounds and differences. We propose avenues for future research and highlight how stronger integration of both research streams could support the development of more powerful approaches to help us navigate toward safe and just futures.
\end{abstract}

\begin{keywords}
resilience, sustainability, transformations, adaptive capacity, regime shifts, adaptive management
\end{keywords}

@@

Social-ecological systems research seeks to understand how social actions influence ecological systems and, vice versa, how dynamics in ecological systems affect social systems [@berkes2000linking].
These systems are intertwined, and their mutual interactions are key for sustained well-being of both ecosystems and people.
Nonlinear changes are of particular concern to sustainability scientists because they tend to be harder to predict, cope with or adapt to, and costly and sometimes impossible to reverse.
Nonlinear changes occur as a consequence of complex social-ecological dynamics responding to disturbances in ways that are not smooth, gradual, or proportional to the magnitude of the disturbance (Fig \ref{fig:fig1}).

Two main strands of sustainability science address nonlinear changes in social-ecological systems, namely the literature on regime shifts [@folke2004] and transformations [@olsson2006shooting].
Although the two concepts share a common origin [@holling1973resilience], they have diverged into different research strands over the past decades (Fig \ref{fig:fig1}).
While regime shifts are large, abrupt hard-to-reverse changes in the function and structure of social-ecological systems [@biggs2018; @folke2004], transformations are often described as involving intentional change that aim to fundamentally shift development of a system into emergent pathways for particular goals, like sustainability and justice [@folke2016; @obrien2025].

The context in which both regime shifts and transformations take place are social-ecological systems.
Social-ecological systems (SES) are intertwined, complex adaptive systems where people interact with biophysical components.
These systems often exhibit nonlinear dynamics with threshold effects, feedback loops, time lags and heterogeneity in system response and resilience [@berkes2000linking; @liu2007a; @berkes2008navigating].
Resilience is the ability of a system to deal with disturbances without permanently losing its function, structure and identity [@folke2016; @holling1973resilience].
Research on the resilience of social-ecological systems emphasizes three important dynamic features: resilience as the capacity to cope with, adapt to, or transform in response to change.
Coping refers to the ability to absorb or resist change to sustain current system dynamics; adaptability is the capacity to learn and combine experience and knowledge to adjust responses and system dynamics; while transformability is the ability to cross thresholds and move the SES into emergent and often unknown development trajectories [@folke2016].

The apparent separation of the two streams of research motivates our review (Fig \ref{fig:fig1}, \ref{fig:cross-ref}).
Similar reviews have already considered the overlaps and distinctions between other concepts, such as transitions and transformations [@loorbach2017].
Both regime shifts and transformations  describe and are applied to large-scale nonlinear changes between states of social-ecological systems and both emerged from research on resilience [@holling1973resilience].
However, these research communities have developed independent conceptual frameworks, methods and modes of application.
By limited engagement with the other, there is a risk for missing opportunities and cross-fertilization between the conceptualizations, thereby reducing the possibility to successfully approach and solve some of the core problems of sustainability science, which by definition, requires understanding social and biophysical dynamics simultaneously.

In particular, dimensions that are central to one framework are at risk of becoming oversimplified in the other.
For example, agency —the capacity of actors to act upon their own choices — is central within transformations research [@olsson2006shooting; @moore2014studying; @benessaiah2021] but less prominent within regime shifts studies, with some exceptions in economics [@li2024].
Conversely, the role of feedbacks or time scales are central in explaining regime shifts [@nyström2012; @folke2004; @rocha2018], but not always assessed in transformations research [@olsson2010building], with some exceptions [@enfors2013].

This lack of dialogue has prevented the emergence of a consensus in terminology, leading to oversimplification, and separate jargon denoting similar characteristics.
For example, the concept of positive tipping points emphasizes desirable transformative changes such as the adoption of electric vehicles [@lenton2025; @lenton2023global], but often omits an analysis of equity or which societal groups will benefit or suffer in this transition, and such changes implies for social-ecological sustainability [@pereira2024equity; @pereira2025].
In contrast, issues related to justice and power have been studied for three decades in transformations research [@pereira2024equity; @blythe2018] highlighting that transformations always involve shifting power relations and entail socially differentiated risks and benefits; and can have unintended consequences [@blythe2018].

The parallel development of similar conceptual frameworks, without scientific dialogue between them, risks reducing the capacity of scientists to test their validity or adequacy in explaining real world problems [@clark2020; @watts2014].
As we will review, some of the criticisms of regime shifts and transformations are lack of comparative case studies that shed light under which conditions these frameworks have explanatory power, or under which conditions are they likely to fail.
An additional risk of multiple untested frameworks is the use of slightly different wording to mean the same thing (e.g. positive tipping points and transformations).
Here, we will review and clarify the nuance on meanings to avoid falling into the jargon trap, or at least make the content more accessible from a pedagogical point of view.
The next section dives on the definitions and history of these concepts, trying to disentangle what are their common grounds and differences.
We then outline current research challenges and areas where we believe the two strands of work could benefit from working together.
Last, we conclude with a short agenda for future research.

\begin{figure*}[h!]
\centering
\includegraphics[width =1\textwidth]{fig1_ARER}
\caption{\textbf{Regime shifts and transformation research.} The potential landscape has inspired heuristic models of regime shifts (A) and transformations (B). A cartoon of the cusp bifurcation for forest and savanna shows that the same system can show semi-linear, abrupt but reversible, and discontinuous change with hysteresis (C). The top 30 papers on regime shifts ranked by relevance are organized on a timeline for regime shifts (D) and transformations (F). There is a lack of cross-fertilization between both strands of literature, with an annual proportion of references lower than 5\% and declining since 2020 (E), Figure \ref{fig:cross-ref} dissagregates the proportion per year. The idea of potential attractors is common to both, but there are differences on how they are studied. This represent an opportunity for cross-fertilization.} 
\label{fig:fig1}
\end{figure*}

@@

Historically, the concept of resilience, as a fundamental characteristic of the regime shifts and transformation frameworks, has its roots in mathematics.
In his seminal work introducing resilience, Holling [@holling1973resilience] suggests minimal models to illustrate how ecosystems can change, lose stability, and shift from one configuration to another, building on the mathematical notion of bifurcations (e.g. [@Poincare:1885ji; @thom1972stabilité] ).
Several concepts such as the basin of attraction (or attractor), tipping points, thresholds, or stability landscapes have precise definitions within bifurcation theory, but are sometimes used as metaphors in later conceptualizations.
A basin of attraction, for example, is the region of parameter space in which all trajectories converge to a common set of points, the so-called attractor.
Attractors can consist of single equilibrium points, but they can also form more complex configurations, such as (a)periodic cycles, long transients, or chaos [@strogatz1994nonlinear].
We will not present formal definitions of these concepts, but instead direct the interested reader to introductory books on the matter [@strogatz1994nonlinear; @sole2011phase; @kuznetsov2023].

In the mathematical context of dynamical systems, resilience has been measured via the size of the basin of attraction [@holling1973resilience; @ives1995; @Clark1975; @krakovská2023], the distance to its boundary, its depth, or the rate of return to equilibrium after a disturbance, among other measures [@Clark1975; @Mitra:2015jb; @Menck:2013jq].
These definitions have enabled the development of early warning signals for the proximity to a tipping point based on fluctuations of the system state [@scheffer2009].
Bifurcation theory, and thus notions of resilience, have been successfully applied to fields other than sustainability science and ecology, including but not limited to language, climate systems, the states of matter, cancer dynamics, finance, social dynamics such as poverty traps, diffusion of beliefs, voting behavior, segregation, the states of mind, or music, to name a few [@sole2011phase; @scheffer2009critical].
Some of these applications precede the work of Holling in ecology [@holling1973resilience], outlining the success of bifurcation theory across the sciences.
However, the evolutionary dynamics of adaptations and transformations are not well defined within the seminal work on bifurcations.
While some formalizations have been used to explore for example the evolution of innovations [@kauffman1987; @kauffman2000; @sole2011phase], it remains an open problem to theorize true generative models that produce agents with behaviors that were not previously coded by the programmer [@kauffman2023].
This problem is common ground to both regime shifts and transformations research.

@@

Regime shifts are abrupt and persistent changes in the structure and function of social-ecological systems [@folke2004; @biggs2018].
They occur when resilience of the current regime is eroded and the system moves from its original configuration into different modes of behavior.
Examples of regime shifts includes the transition from corals to algae-dominated reefs [@hughes2017], eutrophication of lakes [@carpenter2020], the shift from forest to savannahs [@hirota2011; @staver2011], the disappearance of Arctic sea ice [@livina2013], the collapse of the thermohaline circulation [@stocker1997; @stommel1961], or the collapse of cooperation [@rand2014; @nowak2006].
Regime shifts are important because they often affect the flow of ecosystem services that people benefit from [@carpenter2009], and they present a difficult management problem since they can be hard to predict [@biggs2009; @hastings2010] and sometimes impossible to reverse on time scales relevant for management and policy [@biggs2009; @biggs2018].

@@

Regime shifts are typically characterized by feedbacks, tipping points, thresholds, and hysteresis.
Feedbacks are mutually interacting processes that can amplify or dampen responses to drivers [@roe2009].
For example, moisture recycling feedbacks maintain tropical rain forests like the Amazon: more forest transpires more humidity, feeding rainfall that sustains the forest itself[@zemp2017].
In drylands, facilitation feedbacks through shading, soil moisture and nutrient retention are responsible for shifts between grasslands, shrublands and deserts [@maestre2016; @berdugo2022].
In coral reefs, algae can preempt space and increase sediment retention, reducing coral recruitment.
As more reefs shift to algal dominance, large-scale feedbacks such as regional recruitment failure may reinforce local processes, reducing fertilization success and preventing the regeneration of degraded coral reefs [@nyström2012].
Or in social systems, poverty traps are feedbacks mediated by income, education, or disease, that keeps people in poverty despite development efforts [@bowles2016poverty; @banerjee2012poor].
In small-scale fisheries, poverty can lead to overfishing, which in turn forces fishers to use more destructive fishing methods (e.g. dynamite fishing), locking the system in a vicious cycle [@steneck2009; @cinner2009].
While balancing feedbacks stabilize regimes, reinforcing feedbacks drive out-of-equilibrium dynamics [@biggs2018; @roe2009].
Thus, weakening or strengthening of feedbacks can trigger nonlinear transitions.

@@

Tipping points are critical parameter values at which a system changes its long-term qualitative behavior.
They represent bifurcation points or singularities [@strogatz1994nonlinear; @thom1972stabilité].
Speaking of a point implies the assumption of a one-dimensional system with one co-dimension, that is a very simplified version of reality where there is only one state variable that we care about (e.g. one species) and one driver affecting it (e.g. temperature).
Real social-ecological system consist of multiple interacting entities (e.g. species, businesses, cities) that are driven by multiple interacting forces, which are not necessarily reducible to one (co)dimension.
Complex bifurcations (of “co-dimension” greater than one) can arise from the coincidence of multiple drivers [@levin1999fragile; @rocha2015].
When thinking of tipping phenomena in real-life contexts, one should think of tipping lines or a hypersurface, that include all the combination of critical parameters of all potential drivers and state variables, rather than points.
Stochastic variability can further blur the “point” at which tipping happens.
Tipping points and thresholds are often used interchangeably.
However, in mathematics they have a slightly different meaning, the threshold is the separatrix between two basins of attraction, while the tipping point is the critical value at which a bifurcation occurs [@strogatz1994nonlinear] (Fig \ref{fig:fig1}).

Observing and measuring tipping points or thresholds in real life is challenging, leading some to view the concept not useful beyond theoretical work [@hillebrand2020; @kopp2024].
However, scientists have been able to identify some tipping points with different degrees of certainty.
For example, hypoxia and anoxia are regime shifts common in coastal areas where dissolved oxygen falls to levels that are not tolerable for different groups of organisms.
Levels below $2mlO_{2}/L$ gives rise to hypoxia, a condition where many fish die-off, levels below $0.5mlO_{2}/L$ trigger anoxia, a condition that is toxic also to humans [@diaz2008; @breitburg2018].
Tipping points in precipitation have been proposed and confirmed in observational studies for the transitions from forest to savanna ($1500mm$), savanna to forest ($2000mm$), and savanna to deserts ($<500mm$) [@hirota2011; @staver2011; @aleman2020; @verbesselt2016].
Similarly, global thresholds for aridity has been reported for dryland ecosystems (0.54, 0.7 and 0.8) [@berdugo2020], while thresholds for deforestation in rain forest have been proposed (40% for the Amazon, 25% under precautionary principle, currently at 20%) [@nobre2016; @lovejoy2018].
Several Earth subsystems (e.g. the thermohalince circulation, the Greenland ice sheet) can have tipping points driven by temperature.
A recent literature review suggests that six of these subsystems are highly likely to experience tipping dynamics under $2^\circ C$ of warming, with additional four of them tippping under $4^\circ C$ of warming [@armstrongmckay2022].

Hysteresis is the reason why some regime shifts are difficult, sometimes impossible to reverse (Fig \ref{fig:fig1}).
Hysteresis means that tipping point in one direction of the shift is different from the tipping point in the opposite direction.
Thus, additional efforts are necessary to restore social-ecological systems after they have tipped, and in some cases, full recovery may not be achievable at all.
Much research has focused on understanding, preventing and managing regime shifts, and the interventions needed for recovery.
Instead of listing the different strategies that apply to different ecosystems and contexts (see box on the Regime Shifts Data Base), here we summarize an approach that has been successful in managing nonlinear dynamics in the face of uncertainty.

Adaptive management was developed and advocated by early regime shift scholars to encourage monitoring and experimentation, reduce uncertainty, and inform decision making [@holling1978adaptive; @walters1986adaptive].
It is widely applied in fisheries, where monitoring and modeling are combined to inform near-future maximum sustainable yields; projections that are then updated through landing statistics and independent scientific monitoring [@walters1986adaptive].
Other applications are in fire management, where prescribed burns prevent synchronized fires [@peterson2002; @nikolakis2020]; and in lakes, which can recover from eutrophication after manipulating of nutrient, dredging, or biomanipulation of the food web [@carpenter1999; @smith2009; @scheffer1993].
Mathematical models are often used to explore the space of management options, for example to estimate critical nutrient loads in lakes [@kong2016; @janse2008].
In both cases, monitoring and experimentation help managers get insights on relevant feedbacks or how to adapt strategies and decisions to changing social-ecological systems.

@@

Positive tipping points is another recent framework to think about transformations [@lenton2025; @positive2024].
The concept has been widely used in climate sciences to mean conditions and strategies for fast decarbonization, including but not limited to, the adoption of new technologies such as electric vehicles, solar panels, the electrification of grids, or changing diets.
The literature on positive tipping points have been criticized for over-emphasizing technocratic solutions while ignoring core issues of justice and distributions that transformation research have studied for at least three decades [@pereira2024equity].
Who defines what is positive, and positive for whom, are political questions often ignored.
The problem is that the solutions landscape risk to reinforce unjust and colonial practices of exploitation of minorities or less empowered.
It can lead to temporary or localized shifts (e.g. more electric cars) without fundamentally altering the underlying system dynamics, values and norms that reinforced the undesired regime in first place [@milkoreit2022].

@@

::: {#refs}
:::
